Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-07-01 optim.qmd
small editsg

Significant efforts have been devoted to enhancing the functionality of
PP, improving optimisation (e.g. @posse1995, @marie-sainte2010,
@grochowski2011) and extending it's use to visually explore
high-dimensional data. @cook1995grand introduced the PP guided tour,
which monitors the optimisation visually to see the data projections
leading in and out of the optima. An implementation is available in the
R [@R] package `tourr` [@tourr]. @RJ-2021-105 illustrated how to
diagnose optimisation processes, particularly focusing on the guided
tour, and revealed a need to for improved optimisation. While improving
the quality of the optimisation solutions in the tour is essential, it
is also important to be able to view the data projections as the
optimisation progresses. Integrating the guided tour with a global
optimisation algorithm that is efficient in finding the global optimal
and enables viewing of the projected data during the exploration process
is a goal.

@@

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimisation for the guided tour. It is of
interest to assess how quickly and closely the optimiser reaches a gobal
optima, for various PP indexes that may have differing complexities. To
observe the performance of JSO with different types of PP indexes,
metrics are introduced to capture specific properties of the index
including squintability (based on @barnett1981interpreting's squint
angle) and smoothness as introduced in @laa_using_2020. Here we
mathematically define metrics for squintability and smoothness, which is
new for the field. A series of simulation experiments using specially
structured datasets and various PP indexes are conducted to assess the
behaviour of JSO under different settings and its sensitivity to
hyper-parameter choices (number of jellies and maximum number of tries).
The relationship between the JSO performance, hyper-parameter tuning and
properties of PP indexes is analysed to provide guidance for
practitioners using projection pursuit, and developers of new indexes.

The paper is structured as follows. @sec-background introduces the
background of PP guided tour and reviews existing optimisers and index
functions in the literature. @sec-theory details JSO and introduces
metrics that measure different properties of PP indexes. @sec-sim-deets
details two simulation experiments to assess JSO's performance: one
comparing JSO's permformance improvements relative to an existing
optimiser, Creeping Random Search (CRS), and the other studying the
impact of different PP index properties on optimisation performance.
@sec-sim-res presents the results. @sec-conclusion summarises the work
and provides suggestions for future directions.

@@

The JSO implementation involves several key parameters that control its
search process in optimization problems. These parameters are designed
to guide the exploration and exploitation phases of the algorithm. While
the specific implementation details can vary depending on the version of
the algorithm or its application, we focus on two main parameters that
are most relevant to our application: the number of jellyfish and the
maximum number of trials.

@laa_using_2020 has proposed five criteria for assessing projection
pursuit indexes (smoothness, squintability, flexibility, rotation
invariance, and speed). Since not all the properties affects the
execution of the optimisation, we choose to focus here on the first two
properties, namely *smoothness* (@sec-smoothness) and *squintability*
(@sec-squintability), and propose two metrics to evaluate them.

@@

In our context, we suggest to use this parameter $\nu$ as a measure of
the smoothness of the index function by fitting a Gaussian process prior
with Matérn covariance on a dataset generated by random evaluations of
the index function, as done at the initialization stage of the jellyfish
search optimization. There exist several R packages, such as GpGp
[@guinness2021gpgp] or ExaGeoStatR [@abdulah2023large], to fit the
hyperparameters of a GP covariance function on data. In this project, we
make use of the GpGp package.

@@

In the existing literature, it is commonly understood that a large
squint angle implies that the objective function value is close to
optimal even when we are not very close to the perfect view to see the
structure. A small squint angle means that the PP index value improves
substantially only when we are very close to the perfect view. As such,
low squintability implies rapid improvement in the index value when near
the perfect view. For PP, a small squint angle is considered to be
undesirable because it means that the optimiser needs to be very close
to be able to "see" the optima. Thus, it could be difficult for the
optimiser to find the optima. We now propose a mathematical formulation
of this intuition.

@@

```{=html}
<!--
\[We generate random points that is beyond 1.5 projection distance and
interpolate. Then we fit a kernel or use nonlinear least squares.\]
-->
```
It is expected that for a PP index with high squint angle, the
optimization (@eq-optimization) should make substantial progress early
on. Conversely, for a PP index with low squint angle, it might take a
long while for the optimization to make substantial progress, as the
candidate projections would need to be very close to the optimal one for
the structure of the index function to be visible enough to be amenable
to efficient optimization. This observation suggests that the extreme
values of $f'$ (the ones for which $f''=0$, assuming that $f$ is twice
differentiable), and the projection distances for which these values are
attained, should play an important role in the mathematical definition
of squintability. Noticing that $f$ is expected to be a decreasing
function, we propose the following squintability metric:

@@

The JSO performance is compared with the existing Creeping Random Search
(CRS) [@RJ-2021-105; @laa_using_2020] currently used in guided tour, and
explores how JSO behaves with different hyper-parameter (number of
jellies, maximum tries) and data dimension combinations. A second
simulation studies the effect of index properties (smoothness and
squintability), along with JSO hyper-parameters, and data dimension, on
the success rate of the JSO performance.

@@

In this simulation, 50 jellies (`n_jellies`) and a maximum of 50 tries
(`max_tries`) are used. The full data contains 50 simulations, indexed
by the variable `sim`, with the associated `seed`s. Recorded in columns
`basis`, `index_val`, `tries`, `loop` and `time`, are the projection
basis, index value, iteration ID, jellyfish ID, and computational time,
respectively.
<!-- The basis column records every visited basis before -->
<!-- comparing with the current (last) basis. If a visited basis has a -->
<!-- smaller index value than the current (last) one, JSO will keep the -->
<!-- current basis for the next iteration in the search but still records the -->
<!-- visited inferior basis.  -->

In the second experiment, 50 simulations are conducted at each
hyper-parameter combination to analyse its effect on the JSO success
rate: number of jellies (20, 50, and 100), and maximum number of tries
(50 and 100). @fig-success-rate illustrates the calculation based on the
projection found by the above simulation. Here, the optimal index value
is
`r sim_sine_6d_spline_best |> pull(index_val) |> max() |> round(3) -> aa; aa`.
Out of the 50 simulations,
`r sim_sine_6d_spline_best |> filter(index_val >= aa - 0.05) |> nrow() -> bb; bb`
achieved an index value within \[`r aa - 0.05`, `r aa`\], resulting in a
success rate of
`r sim_sine_6d_spline_best |> mutate(a = index_val >= aa - 0.05) |> pull(a) -> cc; round(sum(cc)/length(cc), 3)`.

@@

This paper has presented new metrics to mathematically define desirable
features of PP indexes, squintability and smoothness, and used these to
assess the performance of the new jellyfish search optimiser. The
metrics will be generally useful for characterising PP indexes, and halp
with developing nex indexes.

@@

The JSO performance is affected by the hyper-parameters, with a higher
chance of reaching the global optima when more jellies are used and the
maximum number of tries is increased. However, it comes at a
computational cost, as expected. The performance declines if the
projection dimension increases and if the PP index has low
squintability. The higher the squintability the better chance the JSO
can find the optima. However, interestingly smoothness does not affect
the JSO performance.

The new JSO is integrated with the current implementation of the
projection pursuit guided tour in the `tourr` package, and can be
further examined using PP optimisation diagnostics in the `ferrn`
package. Using the JSO is a little different than the current CRS,
because it runs many paths. The recommended approach is to conduct the
optimisation off-line, and then use the guided tour to follow selected
jellies. The tools for this are available in the `tourr` package, too.

>>>>

Significant efforts have been devoted to enhancing the functionality of
PP, improving optimisation (e.g. @posse1995, @marie-sainte2010,
@grochowski2011) and extending its use to visually explore
high-dimensional data. @cook1995grand introduced the PP guided tour,
which monitors the optimisation visually to see the data projections
leading in and out of the optima. An implementation is available in the
R [@R] package `tourr` [@tourr]. @RJ-2021-105 illustrated how to
diagnose optimisation processes, particularly focusing on the guided
tour, and revealed a need for improved optimisation. While improving
the quality of the optimisation solutions in the tour is essential, it
is also important to be able to view the data projections as the
optimisation progresses. Integrating the guided tour with a global
optimisation algorithm that is efficient in finding the global optimal
and enables viewing of the projected data during the exploration process
is a goal.

@@

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimisation for the guided tour. It is of
interest to assess how quickly and closely the optimiser reaches a gobal
optima, for various PP indexes that may have differing complexities. To
observe the performance of JSO with different types of PP indexes,
metrics are introduced to capture specific properties of the index
including squintability (based on @barnett1981interpreting's squint
angle) and smoothness as introduced in @laa_using_2020. Here we
mathematically define metrics for squintability and smoothness, which is
new for the field. A series of simulation experiments using specially
structured datasets and various PP indexes are conducted to assess the
behaviour of JSO under different settings and its sensitivity to
hyper-parameter choices (number of jellyfishes and maximum number of tries).
The relationship between the JSO performance, hyper-parameter tuning and
properties of PP indexes is analysed to provide guidance for
practitioners using projection pursuit, and developers of new indexes.

The paper is structured as follows. @sec-background introduces the
background of PP guided tour and reviews existing optimisers and index
functions in the literature. @sec-theory details JSO and introduces
metrics that measure different properties of PP indexes. @sec-sim-deets
details two simulation experiments to assess JSO's performance: one
comparing JSO's performance improvements relative to an existing
optimiser, Creeping Random Search (CRS), and the other studying the
impact of different PP index properties on optimisation performance.
@sec-sim-res presents the results. @sec-conclusion summarises the work
and provides suggestions for future directions.

@@

The JSO implementation involves several key parameters that control its
search process in optimization problems. These parameters are designed
to guide the exploration and exploitation phases of the algorithm. While
the specific implementation details can vary depending on the version of
the algorithm or its application, the focus is on two main parameters that
are most relevant to our application: the number of jellyfish and the
maximum number of trials.

@laa_using_2020 has proposed five criteria for assessing projection
pursuit indexes (smoothness, squintability, flexibility, rotation
invariance, and speed). Since not all the properties affects the
execution of the optimisation, the focus here is on the first two
properties, namely *smoothness* (@sec-smoothness) and *squintability*
(@sec-squintability), and propose two metrics to evaluate them.

@@

In this context, the parameter $\nu$ is suggested as a measure of
the smoothness of the index function by fitting a Gaussian process prior
with Matérn covariance on a dataset generated by random evaluations of
the index function, as done at the initialization stage of the jellyfish
search optimization. There exist several R packages, such as GpGp
[@guinness2021gpgp] or ExaGeoStatR [@abdulah2023large], to fit the
hyperparameters of a GP covariance function on data. In this project, we
make use of the GpGp package.

@@

In the existing literature, it is commonly understood that a large
squint angle implies that the objective function value is close to
optimal even when the perfect view to see the structure is far away.
A small squint angle means that the PP index value improves
substantially only when the perfect view is close by. As such,
low squintability implies rapid improvement in the index value when near
the perfect view. For PP, a small squint angle is considered to be
undesirable because it means that the optimiser needs to be very close
to be able to "see" the optima. Thus, it could be difficult for the
optimiser to find the optima. The mathematical formulation
of this intuition is proposed below:

@@

```{=html}
<!--
\[We generate random points that is beyond 1.5 projection distance and
interpolate. Then we fit a kernel or use nonlinear least squares.\]
-->
```
It is expected that for a PP index with high squint angle, the
optimization (@eq-optimization) should make substantial progress early
on. Conversely, for a PP index with low squint angle, it might take a
long while for the optimization to make substantial progress, as the
candidate projections would need to be very close to the optimal one for
the structure of the index function to be visible enough to be amenable
to efficient optimization. This observation suggests that the extreme
values of $f'$ (the ones for which $f''=0$, assuming that $f$ is twice
differentiable), and the projection distances for which these values are
attained, should play an important role in the mathematical definition
of squintability. Noticing that $f$ is expected to be a decreasing
function, the squintability metric is proposed as follows:

@@

The JSO performance is compared with the existing Creeping Random Search
(CRS) [@RJ-2021-105; @laa_using_2020] currently used in guided tour, and
explores how JSO behaves with different hyper-parameter (number of
jellyfishes and maximum tries) and data dimension combinations. A second
simulation studies the effect of index properties (smoothness and
squintability), along with JSO hyper-parameters, and data dimension, on
the success rate of the JSO performance.

@@

In this simulation, 50 jellyfishes (`n_jellies`) and a maximum of 50 tries
(`max_tries`) are used. The full data contains 50 simulations, indexed
by the variable `sim`, with the associated `seed`s. Recorded in columns
`basis`, `index_val`, `tries`, `loop` and `time`, are the projection
basis, index value, iteration ID, jellyfish ID, and computational time,
respectively.
<!-- The basis column records every visited basis before -->
<!-- comparing with the current (last) basis. If a visited basis has a -->
<!-- smaller index value than the current (last) one, JSO will keep the -->
<!-- current basis for the next iteration in the search but still records the -->
<!-- visited inferior basis.  -->

In the second experiment, 50 simulations are conducted at each
hyper-parameter combination to analyse its effect on the JSO success
rate: number of jellyfishes (20, 50, and 100), and maximum number of tries
(50 and 100). @fig-success-rate illustrates the calculation based on the
projection found by the above simulation. Here, the optimal index value
is
`r sim_sine_6d_spline_best |> pull(index_val) |> max() |> round(3) -> aa; aa`.
Out of the 50 simulations,
`r sim_sine_6d_spline_best |> filter(index_val >= aa - 0.05) |> nrow() -> bb; bb`
achieved an index value within \[`r aa - 0.05`, `r aa`\], resulting in a
success rate of
`r sim_sine_6d_spline_best |> mutate(a = index_val >= aa - 0.05) |> pull(a) -> cc; round(sum(cc)/length(cc), 3)`.

@@

This paper has presented new metrics to mathematically define desirable
features of PP indexes, squintability and smoothness, and used these to
assess the performance of the new jellyfish search optimiser. The
metrics will be generally useful for characterising PP indexes, and help
with developing new indexes.

@@

The JSO performance is affected by the hyper-parameters, with a higher
chance of reaching the global optima when more jellyfishes are used and the
maximum number of tries is increased. However, it comes at a
computational cost, as expected. The performance declines if the
projection dimension increases and if the PP index has low
squintability. The higher the squintability the better chance the JSO
can find the optima. However, interestingly smoothness does not affect
the JSO performance.

The new JSO is integrated with the current implementation of the
projection pursuit guided tour in the `tourr` package, and can be
further examined using PP optimisation diagnostics in the `ferrn`
package. Using the JSO is a little different than the current CRS,
because it runs many paths. The recommended approach is to conduct the
optimisation off-line, and then use the guided tour to follow selected
jellyfish. The tools for this are available in the `tourr` package, too.

### 2024-07-01 optim.tex
small editsg

\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criteria function known as the PP index, to explore high-dimensional
data by revealing interesting projections. The optimisation in PP can be
non-trivial, involving non-smooth functions and optima with small a
``squint angle'', detectable only from close proximity. To address these
challenges, this study investigates the performance of a recently
introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for
optimising PP indexes. The performance of JSO for visualising data is
evaluated across various hyper-parameter settings and compared with
existing optimisers. Additionally, this work proposes novel methods to
quantify two properties of the PP index -- smoothness and squintability
-- that capture the complexities inherent in PP optimisation problems.
These two metrics are evaluated with JSO hyper-parameters to determine
their effects on JSO success rate. The JSO algorithm has been
implemented in the \texttt{tourr} package and functions to calculate
smoothness and squintability of indexes are available in the
\texttt{ferrn} package.
\end{abstract}

@@

Significant efforts have been devoted to enhancing the functionality of
PP, improving optimisation (e.g. \citet{posse1995},
\citet{marie-sainte2010}, \citet{grochowski2011}) and extending it's use
to visually explore high-dimensional data. \citet{cook1995grand}
introduced the PP guided tour, which monitors the optimisation visually
to see the data projections leading in and out of the optima. An
implementation is available in the R \citep{R} package \texttt{tourr}
\citep{tourr}. \citet{RJ-2021-105} illustrated how to diagnose
optimisation processes, particularly focusing on the guided tour, and
revealed a need to for improved optimisation. While improving the
quality of the optimisation solutions in the tour is essential, it is
also important to be able to view the data projections as the
optimisation progresses. Integrating the guided tour with a global
optimisation algorithm that is efficient in finding the global optimal
and enables viewing of the projected data during the exploration process
is a goal.

@@

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimisation for the guided tour. It is of
interest to assess how quickly and closely the optimiser reaches a gobal
optima, for various PP indexes that may have differing complexities. To
observe the performance of JSO with different types of PP indexes,
metrics are introduced to capture specific properties of the index
including squintability (based on \citet{barnett1981interpreting}'s
squint angle) and smoothness as introduced in \citet{laa_using_2020}.
Here we mathematically define metrics for squintability and smoothness,
which is new for the field. A series of simulation experiments using
specially structured datasets and various PP indexes are conducted to
assess the behaviour of JSO under different settings and its sensitivity
to hyper-parameter choices (number of jellies and maximum number of
tries). The relationship between the JSO performance, hyper-parameter
tuning and properties of PP indexes is analysed to provide guidance for
practitioners using projection pursuit, and developers of new indexes.

The paper is structured as follows. Section~\ref{sec-background}
introduces the background of PP guided tour and reviews existing
optimisers and index functions in the literature.
Section~\ref{sec-theory} details JSO and introduces metrics that measure
different properties of PP indexes. Section~\ref{sec-sim-deets} details
two simulation experiments to assess JSO's performance: one comparing
JSO's permformance improvements relative to an existing optimiser,
Creeping Random Search (CRS), and the other studying the impact of
different PP index properties on optimisation performance.
Section~\ref{sec-sim-res} presents the results.
Section~\ref{sec-conclusion} summarises the work and provides
suggestions for future directions.

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, coltitle=black, colback=white, colbacktitle=quarto-callout-note-color!10!white, left=2mm, leftrule=.75mm, titlerule=0mm, rightrule=.15mm, toptitle=1mm, opacityback=0, arc=.35mm, bottomrule=.15mm, bottomtitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, toprule=.15mm, opacitybacktitle=0.6, breakable]

@@

The JSO implementation involves several key parameters that control its
search process in optimization problems. These parameters are designed
to guide the exploration and exploitation phases of the algorithm. While
the specific implementation details can vary depending on the version of
the algorithm or its application, we focus on two main parameters that
are most relevant to our application: the number of jellyfish and the
maximum number of trials.

\citet{laa_using_2020} has proposed five criteria for assessing
projection pursuit indexes (smoothness, squintability, flexibility,
rotation invariance, and speed). Since not all the properties affects
the execution of the optimisation, we choose to focus here on the first
two properties, namely \emph{smoothness} (Section~\ref{sec-smoothness})
and \emph{squintability} (Section~\ref{sec-squintability}), and propose
two metrics to evaluate them.

@@

In our context, we suggest to use this parameter \(\nu\) as a measure of
the smoothness of the index function by fitting a Gaussian process prior
with Matérn covariance on a dataset generated by random evaluations of
the index function, as done at the initialization stage of the jellyfish
search optimization. There exist several R packages, such as GpGp
\citep{guinness2021gpgp} or ExaGeoStatR \citep{abdulah2023large}, to fit
the hyperparameters of a GP covariance function on data. In this
project, we make use of the GpGp package.

@@

In the existing literature, it is commonly understood that a large
squint angle implies that the objective function value is close to
optimal even when we are not very close to the perfect view to see the
structure. A small squint angle means that the PP index value improves
substantially only when we are very close to the perfect view. As such,
low squintability implies rapid improvement in the index value when near
the perfect view. For PP, a small squint angle is considered to be
undesirable because it means that the optimiser needs to be very close
to be able to ``see'' the optima. Thus, it could be difficult for the
optimiser to find the optima. We now propose a mathematical formulation
of this intuition.

@@

It is expected that for a PP index with high squint angle, the
optimization (\ref{eq-optimization}) should make substantial progress
early on. Conversely, for a PP index with low squint angle, it might
take a long while for the optimization to make substantial progress, as
the candidate projections would need to be very close to the optimal one
for the structure of the index function to be visible enough to be
amenable to efficient optimization. This observation suggests that the
extreme values of \(f'\) (the ones for which \(f''=0\), assuming that
\(f\) is twice differentiable), and the projection distances for which
these values are attained, should play an important role in the
mathematical definition of squintability. Noticing that \(f\) is
expected to be a decreasing function, we propose the following
squintability metric:

@@

The JSO performance is compared with the existing Creeping Random Search
(CRS) \citep{RJ-2021-105, laa_using_2020} currently used in guided tour,
and explores how JSO behaves with different hyper-parameter (number of
jellies, maximum tries) and data dimension combinations. A second
simulation studies the effect of index properties (smoothness and
squintability), along with JSO hyper-parameters, and data dimension, on
the success rate of the JSO performance.

@@

In this simulation, 50 jellies (\texttt{n\_jellies}) and a maximum of 50
tries (\texttt{max\_tries}) are used. The full data contains 50
simulations, indexed by the variable \texttt{sim}, with the associated
\texttt{seed}s. Recorded in columns \texttt{basis}, \texttt{index\_val},
\texttt{tries}, \texttt{loop} and \texttt{time}, are the projection
basis, index value, iteration ID, jellyfish ID, and computational time,
respectively.

In the second experiment, 50 simulations are conducted at each
hyper-parameter combination to analyse its effect on the JSO success
rate: number of jellies (20, 50, and 100), and maximum number of tries
(50 and 100). Figure~\ref{fig-success-rate} illustrates the calculation
based on the projection found by the above simulation. Here, the optimal
index value is 0.999. Out of the 50 simulations, 44 achieved an index
value within {[}0.949, 0.999{]}, resulting in a success rate of 0.88.

@@

The JSO performance is affected by the hyper-parameters, with a higher
chance of reaching the global optima when more jellies are used and the
maximum number of tries is increased. However, it comes at a
computational cost, as expected. The performance declines if the
projection dimension increases and if the PP index has low
squintability. The higher the squintability the better chance the JSO
can find the optima. However, interestingly smoothness does not affect
the JSO performance.

The new JSO is integrated with the current implementation of the
projection pursuit guided tour in the \texttt{tourr} package, and can be
further examined using PP optimisation diagnostics in the \texttt{ferrn}
package. Using the JSO is a little different than the current CRS,
because it runs many paths. The recommended approach is to conduct the
optimisation off-line, and then use the guided tour to follow selected
jellies. The tools for this are available in the \texttt{tourr} package,
too.

>>>>

\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criteria function known as the PP index, to explore high-dimensional
data by revealing interesting projections. The optimisation in PP can be
non-trivial, involving non-smooth functions and optima with a small
``squint angle'', detectable only from close proximity. To address these
challenges, this study investigates the performance of a recently
introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for
optimising PP indexes. The performance of JSO for visualising data is
evaluated across various hyper-parameter settings and compared with
existing optimisers. Additionally, this work proposes novel methods to
quantify two properties of the PP index --~smoothness and
squintability~-- that capture the complexities inherent in PP
optimisation problems. These two metrics are evaluated along with JSO
hyper-parameters to determine their effects on JSO success rate. Our
numerical results confirm the positive impact of these metrics on the
JSO success rate, with squintability being the most significant. The JSO
algorithm has been implemented in the \texttt{tourr} package and
functions to calculate smoothness and squintability of indexes are
available in the \texttt{ferrn} package.
\end{abstract}

@@

Significant efforts have been devoted to enhancing the functionality of
PP, improving optimisation (e.g. \citet{posse1995},
\citet{marie-sainte2010}, \citet{grochowski2011}) and extending its use
to visually explore high-dimensional data. \citet{cook1995grand}
introduced the PP guided tour, which monitors the optimisation visually
to see the data projections leading in and out of the optima. An
implementation is available in the R \citep{R} package \texttt{tourr}
\citep{tourr}. \citet{RJ-2021-105} illustrated how to diagnose
optimisation processes, particularly focusing on the guided tour, and
revealed a need for improved optimisation. While improving the quality
of the optimisation solutions in the tour is essential, it is also
important to be able to view the data projections as the optimisation
progresses. Integrating the guided tour with a global optimisation
algorithm that is efficient in finding the global optimal and enables
viewing of the projected data during the exploration process is a goal.

@@

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimisation for the guided tour. It is of
interest to assess how quickly and closely the optimiser reaches a gobal
optima, for various PP indexes that may have differing complexities. To
observe the performance of JSO with different types of PP indexes,
metrics are introduced to capture specific properties of the index
including squintability (based on \citet{barnett1981interpreting}'s
squint angle) and smoothness as introduced in \citet{laa_using_2020}.
Here we mathematically define metrics for squintability and smoothness,
which is new for the field. A series of simulation experiments using
specially structured datasets and various PP indexes are conducted to
assess the behaviour of JSO under different settings and its sensitivity
to hyper-parameter choices (number of jellyfishes and maximum number of
tries). The relationship between the JSO performance, hyper-parameter
tuning and properties of PP indexes is analysed to provide guidance for
practitioners using projection pursuit, and developers of new indexes.

The paper is structured as follows. Section~\ref{sec-background}
introduces the background of PP guided tour and reviews existing
optimisers and index functions in the literature.
Section~\ref{sec-theory} details JSO and introduces metrics that measure
different properties of PP indexes. Section~\ref{sec-sim-deets} details
two simulation experiments to assess JSO's performance: one comparing
JSO's performance improvements relative to an existing optimiser,
Creeping Random Search (CRS), and the other studying the impact of
different PP index properties on optimisation performance.
Section~\ref{sec-sim-res} presents the results.
Section~\ref{sec-conclusion} summarises the work and provides
suggestions for future directions.

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, rightrule=.15mm, breakable, opacityback=0, toptitle=1mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, bottomrule=.15mm, arc=.35mm, toprule=.15mm, bottomtitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, title={Algorithm: Jellyfish Optimizer Pseudo Code}, leftrule=.75mm, left=2mm]

@@

The JSO implementation involves several key parameters that control its
search process in optimization problems. These parameters are designed
to guide the exploration and exploitation phases of the algorithm. While
the specific implementation details can vary depending on the version of
the algorithm or its application, the focus is on two main parameters
that are most relevant to our application: the number of jellyfish and
the maximum number of trials.

\citet{laa_using_2020} has proposed five criteria for assessing
projection pursuit indexes (smoothness, squintability, flexibility,
rotation invariance, and speed). Since not all the properties affects
the execution of the optimisation, the focus here is on the first two
properties, namely \emph{smoothness} (Section~\ref{sec-smoothness}) and
\emph{squintability} (Section~\ref{sec-squintability}), and propose two
metrics to evaluate them.

@@

In this context, the parameter \(\nu\) is suggested as a measure of the
smoothness of the index function by fitting a Gaussian process prior
with Matérn covariance on a dataset generated by random evaluations of
the index function, as done at the initialization stage of the jellyfish
search optimization. There exist several R packages, such as GpGp
\citep{guinness2021gpgp} or ExaGeoStatR \citep{abdulah2023large}, to fit
the hyperparameters of a GP covariance function on data. In this
project, we make use of the GpGp package.

@@

In the existing literature, it is commonly understood that a large
squint angle implies that the objective function value is close to
optimal even when the perfect view to see the structure is far away. A
small squint angle means that the PP index value improves substantially
only when the perfect view is close by. As such, low squintability
implies rapid improvement in the index value when near the perfect view.
For PP, a small squint angle is considered to be undesirable because it
means that the optimiser needs to be very close to be able to ``see''
the optima. Thus, it could be difficult for the optimiser to find the
optima. The mathematical formulation of this intuition is proposed
below:

@@

It is expected that for a PP index with high squint angle, the
optimization (\ref{eq-optimization}) should make substantial progress
early on. Conversely, for a PP index with low squint angle, it might
take a long while for the optimization to make substantial progress, as
the candidate projections would need to be very close to the optimal one
for the structure of the index function to be visible enough to be
amenable to efficient optimization. This observation suggests that the
extreme values of \(f'\) (the ones for which \(f''=0\), assuming that
\(f\) is twice differentiable), and the projection distances for which
these values are attained, should play an important role in the
mathematical definition of squintability. Noticing that \(f\) is
expected to be a decreasing function, the squintability metric is
proposed as follows:

@@

The JSO performance is compared with the existing Creeping Random Search
(CRS) \citep{RJ-2021-105, laa_using_2020} currently used in guided tour,
and explores how JSO behaves with different hyper-parameter (number of
jellyfishes and maximum tries) and data dimension combinations. A second
simulation studies the effect of index properties (smoothness and
squintability), along with JSO hyper-parameters, and data dimension, on
the success rate of the JSO performance.

@@

In this simulation, 50 jellyfishes (\texttt{n\_jellies}) and a maximum
of 50 tries (\texttt{max\_tries}) are used. The full data contains 50
simulations, indexed by the variable \texttt{sim}, with the associated
\texttt{seed}s. Recorded in columns \texttt{basis}, \texttt{index\_val},
\texttt{tries}, \texttt{loop} and \texttt{time}, are the projection
basis, index value, iteration ID, jellyfish ID, and computational time,
respectively.

In the second experiment, 50 simulations are conducted at each
hyper-parameter combination to analyse its effect on the JSO success
rate: number of jellyfishes (20, 50, and 100), and maximum number of
tries (50 and 100). Figure~\ref{fig-success-rate} illustrates the
calculation based on the projection found by the above simulation. Here,
the optimal index value is 0.999. Out of the 50 simulations, 44 achieved
an index value within {[}0.949, 0.999{]}, resulting in a success rate of
0.88.

@@

The JSO performance is affected by the hyper-parameters, with a higher
chance of reaching the global optima when more jellyfishes are used and
the maximum number of tries is increased. However, it comes at a
computational cost, as expected. The performance declines if the
projection dimension increases and if the PP index has low
squintability. The higher the squintability the better chance the JSO
can find the optima. However, interestingly smoothness does not affect
the JSO performance.

The new JSO is integrated with the current implementation of the
projection pursuit guided tour in the \texttt{tourr} package, and can be
further examined using PP optimisation diagnostics in the \texttt{ferrn}
package. Using the JSO is a little different than the current CRS,
because it runs many paths. The recommended approach is to conduct the
optimisation off-line, and then use the guided tour to follow selected
jellyfish. The tools for this are available in the \texttt{tourr}
package, too.
